---
title: "PA_1 Template"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```

###Loading packages
```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
```

###Load the data and (re)set wd
```{r, echo = TRUE}
start_dir <- getwd()
setwd("../Reproducible Research Week 2/repdata_data_activity")
files <- list.files()
activity_data <- read.csv2(files[1], 
                           sep = ",", 
                           header = TRUE, 
                           stringsAsFactors = FALSE)
setwd(start_dir)
```

###Prepocessing data
```{r, echo=TRUE}
activity_data$date <- as.Date(activity_data$date)
```


#What is mean total number of steps taken per day?
First group data per date to get the total number of steps taken per day: 
```{r, echo=TRUE}
daily_steps <- activity_data %>% 
                group_by(date) %>% 
                summarise(steps_per_day = sum(steps, na.rm = TRUE))
```

Plot a histogram of the total number of steps taken each day:
```{r, echo=TRUE, message=FALSE, warning=FALSE}
plot_total_steps <- daily_steps %>% ggplot(aes(x = steps_per_day)) + 
        geom_histogram() + 
        ggtitle("Histogram of total number of steps per day") + 
        xlab("Total steps per day") +
        ylab("Frequency")
print(plot_total_steps)
```

Calculate and report the mean and median number of steps taken each day:
```{r, echo=TRUE}
mean_steps_day <- mean(daily_steps$steps_per_day, na.rm = TRUE)
median_steps_day <- median(daily_steps$steps_per_day, na.rm = TRUE)
```
Mean of steps taken each day:
```{r, echo=TRUE}

print(mean_steps_day)
```
Median of steps taken each day:
```{r, echo=TRUE}

print(median_steps_day)
```

#What is the average daily activity pattern?

Group data by average amount of steps taken per interval:
```{r, echo=TRUE}
avg_steps_interval <- activity_data %>% 
        group_by(interval) %>% 
        summarise(mean_steps = mean(steps, na.rm = TRUE))
```

Then plot a time series graph to show the daily pattern:
```{r, echo=TRUE}
plot_avg_steps_interval <- avg_steps_interval %>%
        ggplot(aes( x = interval, y = mean_steps)) + 
        geom_line() +
        ggtitle("Average number of steps per 5-minute interval") +
        xlab("5-minute interval") +
        ylab("Average number of steps")
print(plot_avg_steps_interval)
```

Interval with highest average number of steps:
```{r, echo=TRUE}
highest_avg_interval <- avg_steps_interval %>% filter(mean_steps == max(mean_steps, na.rm = TRUE))
print(highest_avg_interval)
```

#Imputing missing values
Amount of missing values in dataset:
```{r, echo=TRUE}
n_na <- sum(is.na(activity_data))
print(n_na)
```

Appearance of NA's
```{r, echo=TRUE}
missing_days <- subset(activity_data, is.na(activity_data$steps))
missing_days <- unique(missing_days$date)
print(missing_days)
```

NA's occur on 8 specific days. Subject likely forgot to put on the activity tracker.
I chose to exclude the missing values from the data. Missing record are removed from the (new) data set:
```{r, echo=TRUE}
activity_data_no_na <- subset(activity_data, !is.na(activity_data$steps))
```


Group data without NA's per date to get the total number of steps taken per day:
```{r, echo=TRUE}
daily_steps_no_na <- activity_data_no_na %>% 
                        group_by(date) %>% 
                        summarise(steps_per_day = sum(steps, na.rm = TRUE))
```


Plot a histogram of total number of steps taken each day after removing missing values:
```{r, echo=TRUE}
plot_total_steps_no_na <- daily_steps_no_na %>% 
        ggplot(aes(x = steps_per_day)) + geom_histogram() + 
        ggtitle("Histogram of total number of steps per day (no NA's)") + 
        xlab("Total steps per day") +
        ylab("Frequency")
print(plot_total_steps_no_na)
```


Calculate and report the mean and median number of steps taken each day with & without missing values:
```{r, echo=TRUE}
mean_steps_no_na <- mean(daily_steps_no_na$steps_per_day, na.rm = TRUE)
median_steps_no_na <- median(daily_steps_no_na$steps_per_day, na.rm = TRUE)
overview <- data.frame(with_na = c(mean_steps_day, median_steps_day), 
                       no_na = c(mean_steps_no_na, median_steps_no_na), 
                       row.names = c("mean", "median"))
print(overview)
```
#Are there differences in activity patterns between weekdays and weekends?

Add variable for type of day of specified date (week- or weekendday) to dataset:
```{r, echo=TRUE}
activity_data_no_na$weekday <- weekdays(activity_data_no_na$date)
activity_data_no_na$day_type <- sapply(activity_data_no_na$weekday, function(x) {
        if (x == "zaterdag" | x == "zondag")
                {y <- "weekend"} else
                {y <- "weekday"}
                y
                })
```

Construct a panel plot containing time series plots of the 5-minute intervals and the average number of steps averaged across all weekdays and weekend days:
```{r, echo=TRUE}
plot_steps_per_interval_type_day <- activity_data_no_na %>% 
        group_by(day_type, interval) %>% 
        summarise(mean_steps = mean(steps, na.rm = TRUE)) %>%
        ggplot(aes( x = interval, y = mean_steps, color = day_type)) + 
        geom_line() +
        ggtitle("Average number of steps per 5-minute interval by type of day") +
        xlab("5-minute interval") +
        ylab("Average number steps") +
        facet_wrap(~day_type, ncol = 1, nrow =2)
print(plot_steps_per_interval_type_day)
```



